Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2014-10-12 algorithm.tex
AA: implementati i suggerimenti sull'espressione inglese del reviewer II

By a slightly abuse of notation, considering a subset $I$ of the configuration space, we will denote with $\mathcal{P}(I,b)$ the non-dominated feasible configurations in $I$, i.e.
	\begin{align}
		\mathcal{P}(I,b)=
		\left\{ \mathbf{c} \in I \cap \mathcal{C}^*(S) | \nexists \ \mathbf{c}' \in I \cap \mathcal{C}^*(S), \mathbf{c}' \succ \mathbf{c} \right\}
	\end{align}
Since in each run of the algorithm we will consider only one benchmark application $b\in\mathcal{B}$, we will omit it, replacing, for example, $\mathcal{P}(I,b)$ with  $\mathcal{P}(I)$.

@@

For $era_{i}$ with $i>0$ the following operations must be performed.
\begin{enumerate}
\item \label{pers02.enu:K}
In each era $i$, a set $test_i$ of $K$ configurations will be evaluated. We call them \emph{test configurations} and selecting them is exactly the goal of this step.
We select only \emph{new} configurations, i.e. the ones that have not been evaluated in past eras. In other words, $test_i$ must not contain $\bigcup_{j=0}^{i-1}test_{j}$.
Since all regions $R$ in the configuration space partition $\mathcal{R}_{i}$ must be properly explored, we have to distribute the test configurations among all these regions. For this reason, $\frac{K}{\left|\mathcal{R}_{i}\right|}$ configurations are randomly extracted from the new configurations of each region $R\in\mathcal{R}_{i}$. We denote the set of the selected configurations with $test_{R,i}$ and, obviously, $\bigcup_{R\in\mathcal{R}_{i}} test_{R,i} = test_i$.
 In case the new configurations in $R$ are less then $\frac{K}{\left|\mathcal{R}_{i}\right|}$, all of them are selected. In this case $test_{R,i}=R\setminus\bigcup_{j=0}^{i-1}test_{j}$, where  $\bigcup_{j=0}^{i-1}test_{j}$ is the set of the configurations already in the past.\footnote{Note that if such regions exist, the total number of evaluations in era $i$ will be less than $K$}
\item All configurations on $test_{i}$ are simulated.
\item The Pareto set $\mathscr{P}_{i}$ for $era_{i}$ is defined as the set of non-dominated configurations among the ones evaluated so far, i.e.
	\[
	\mathscr{P}_{i} \triangleq
	\mathscr{P}\left(\bigcup_{j\leq i}test_j\right)=
	\mathscr{P}\left(test_{i}\cup\mathscr{P}_{i-1}\right)
	\]
where the last equality can easily be proved.

@@

\item The goal of this step is preparing the regions that will compose the configuration space partition of the next era, i.e. $\mathscr{R}_{i+1}$. The following operations are performed:
	\begin{enumerate}
		\item Each high innovation region $R\in\mathscr{R}_{i,h}$ is split obtaining $\psi(R)$, as stated in def \defR{splitting-a-region}. We denote with $\psi\left(\mathscr{R}_{i,h}\right)$ the set of regions obtained as explained.
		\item Low innovation regions are left unchanged.
		\item No innovation regions are merged, two by two. To do so, pairs of no innovation contiguous regions $(R_1,R_2)$ are selected. We denote the set of these pairs as $\mathscr{M}_i$. This set is constructed such that pairs are independent, i.e. there cannot be two of them sharing a region. In other words, taking whatever $(R_1,R_2),(R_3,R_4) \in \mathscr{M}_i$, $R_i \neq R_j, \forall i,j\in\lbrace1,2,3,4\rbrace, i\neq j$. 
We merge each pair of these regions, i.e. $\forall (R_1,R_2)\in\mathscr{M}_i$ we calculate $\mu(R_1,R_2)$, according to \defR{merging-regions}.
We stress that the regions that belong to each selected pair must be contiguous, otherwise it is impossible to merge them. We denote with $\mu(\mathscr{M}_i)$ the set of the merged regions, i.e. $\mu(\mathscr{M}_i) = \bigcup_{(R_1,R_2)\in\mathscr{M}_i} \mu(R_1,R_2)$
	\end{enumerate}

>>>>

By a slightly abuse of notation, considering a subset $I$ of the configuration space, we will use  $\mathcal{P}(I,b)$ to denote the non-dominated feasible configurations in $I$, i.e.
	\begin{align}
		\mathcal{P}(I,b)=
		\left\{ \mathbf{c} \in I \cap \mathcal{C}^*(S) | \nexists \ \mathbf{c}' \in I \cap \mathcal{C}^*(S), \mathbf{c}' \succ \mathbf{c} \right\}
	\end{align}
Since in each run of the algorithm we will consider only one benchmark application $b\in\mathcal{B}$, we will omit it, replacing, for example, $\mathcal{P}(I,b)$ with  $\mathcal{P}(I)$.

@@

For $era_{i}$ with $i>0$ the following operations must be performed.
\begin{enumerate}
\item \label{pers02.enu:K}
In each era $i$, a set $test_i$ of $K$ configurations will be evaluated. We call them \emph{test configurations} and selecting them is exactly the goal of this step.
We select only \emph{new} configurations, i.e. the ones that have not been evaluated in past eras. In other words, $test_i$ must not contain $\bigcup_{j=0}^{i-1}test_{j}$.
Since all regions $R$ in the configuration space partition $\mathcal{R}_{i}$ must be properly explored, we have to distribute the test configurations among all these regions. For this reason, $\frac{K}{\left|\mathcal{R}_{i}\right|}$ configurations are randomly extracted from the new configurations of each region $R\in\mathcal{R}_{i}$. We use $test_{R,i}$ to denote the set of the selected configurations and, obviously, $\bigcup_{R\in\mathcal{R}_{i}} test_{R,i} = test_i$.
 In case the new configurations in $R$ are less then $\frac{K}{\left|\mathcal{R}_{i}\right|}$, all of them are selected. In this case $test_{R,i}=R\setminus\bigcup_{j=0}^{i-1}test_{j}$, where  $\bigcup_{j=0}^{i-1}test_{j}$ is the set of the configurations already in the past.\footnote{Note that if such regions exist, the total number of evaluations in era $i$ will be less than $K$}
\item All configurations on $test_{i}$ are simulated.
\item The Pareto set $\mathscr{P}_{i}$ for $era_{i}$ is defined as the set of non-dominated configurations among the ones evaluated so far, i.e.
	\[
	\mathscr{P}_{i} \triangleq
	\mathscr{P}\left(\bigcup_{j\leq i}test_j\right)=
	\mathscr{P}\left(test_{i}\cup\mathscr{P}_{i-1}\right)
	\]
where the last equality can easily be proved.

@@

\item The goal of this step is preparing the regions that will compose the configuration space partition of the next era, i.e. $\mathscr{R}_{i+1}$. The following operations are performed:
	\begin{enumerate}
		\item Each high innovation region $R\in\mathscr{R}_{i,h}$ is split to obtain $\psi(R)$, as stated in \defR{splitting-a-region}. We use $\psi\left(\mathscr{R}_{i,h}\right)$ to denote the set of regions obtained as explained.
		\item Low innovation regions are left unchanged.
		\item No innovation regions are merged, two by two. To do so, pairs of no innovation contiguous regions $(R_1,R_2)$ are selected. We use $\mathscr{M}_i$ to represent the set of these pairs. This set is constructed such that pairs are independent, i.e. there cannot be two of them sharing a region. In other words, taking whatever $(R_1,R_2),(R_3,R_4) \in \mathscr{M}_i$, $R_i \neq R_j, \forall i,j\in\lbrace1,2,3,4\rbrace, i\neq j$. 
We merge each pair of these regions, i.e. $\forall (R_1,R_2)\in\mathscr{M}_i$ we calculate $\mu(R_1,R_2)$, according to \defR{merging-regions}.
We stress that the regions that belong to each selected pair must be contiguous, otherwise it is impossible to merge them. We use $\mu(\mathscr{M}_i)$ to denote the set of the merged regions, i.e. $\mu(\mathscr{M}_i) = \bigcup_{(R_1,R_2)\in\mathscr{M}_i} \mu(R_1,R_2)$
	\end{enumerate}

### 2014-10-12 concepts.tex
AA: implementati i suggerimenti sull'espressione inglese del reviewer II

% A PROPOSITO DI QUESTO DUBBIO:
%\comment{AA: Mi sembra solo un cambio di notazione. DP: c'era un On the contrary troppo aggressivio, ho cercato di alleggerire la cosa, ma dai un'occhiata }
% HO CANCELLATO TUTTA QUESTA PARTE.
% For each point of the Pareto front we can imagine the corresponding configuration as a point of $P_{1}\times\dots\times P_{n}$ space. We will also refer to these points as ``parameter space representation'' of the Pareto front. Figure~\ref{fig:psos} shows the relation between the objective space (containing the Pareto front) and its parameter space representation. In particular in this work we are interested in the relation between the evolution of the points in the two spaces, e.g. how some properties of the Pareto front points can reflect some features of the corresponding points in the parameter space.

\begin{figure}[t]
\label{fig:psos}
\center
\includegraphics[width=0.5\columnwidth]{img/Pareto_set_and_front}
\caption{Relation between Pareto-set and Pareto front.}
\end{figure}

@@

The problem is that, in complex scenarios, not all the relations of this
kind are known in advance. Some dependency may be hidden
to the designer.
Therefore, as stated in section \ref{sec:Related-work},
setting up an exploration trying to take into account as much dependencies
as possible is a hard task that is destined to be incompletely
accomplished. We need a methodology to ``automatically'' recognize
interesting or uninteresting regions without requiring a priori knowledge
to the designer, that is what our algorithm does.

@@

%TODO: sync name of values with formulation
\subsection{Formal definitions}
We start by defining the parameter interval and the concept of interval contiguity. Using these definition we construct the region and define the region contiguity. Then we define the operations of splitting and merging regions. Finally, we provide the definition of separation.

@@

We consider only parameters $p_{i}$ with ordering, i.e. such
that $\forall a,b\in V_{i}$ it is possible
to say $a<b$ or $a=b$ or $a>b$. 
The following definition of interval contiguity will be the base of the concept of region contiguity, used to define merging and splitting operations.
\begin{definition}[Contiguous intervals]
\label{pers02.def:Contiguous-intervals}
Two intervals are contiguous iff they do not overlap and merging them results in a new interval.More formally, let $\left[a_{i}..b_{i}\right]$ and $\left[x_{i}..y_{i}\right]$
be two $p_{i}$-intervals. They are said to be contiguous (see \figR{interval_and_region}(a) ) iff
\begin{align}\begin{array}{l}
	\left[a_{i}..b_{i}\right]\cap\left[x_{i}..y_{i}\right]=\emptyset \mbox{      and} \\
	\left[a_{i}..b_{i}\right]\cup\left[x_{i}..y_{i}\right]$ is a $p_{i}$-
interval or $\left[x_{i}..y_{i}\right]\cup\left[a_{i}..b_{i}\right]
\end{array}\end{align}
\end{definition}

@@

As already explain, we will split interesting regions and merge uninteresting ones. The region splitting and merging operations are defined starting from the corresponding operations defined on intervals.
\begin{definition}[Splitting an interval]
\defL{splitting-a-region}
Given a $p_i$ interval $\left[a_{i}..b_{i}\right]$ with more than one element, i.e. $\left|\left[a_{i}..b_{i}\right]\right|>1$, the splitting operation produces as result two contiguous intervals $\left[a_{i}..c_{i}\right],\left[d_{i}..b_{i}\right]$
 such that
	\begin{align}
		\begin{cases}
		\left[a_{i}..c_{i}\right]\cap\left[d_{i}..b_{i}\right] & =\emptyset\\
		\left[a_{i}..c_{i}\right]\cup\left[d_{i}..b_{i}\right] & =\left[a_{i}..b_{i}\right]\\
		\left|\left[a_{i}..c_{i}\right]\right| & =\left\lceil \frac{\left|\left[a_{i}..b_{i}\right]\right|}{2}\right\rceil 
		\end{cases}
	\end{align}

We define the split operator $\psi$ as the one that, applied to the intervals, gives the result of the split operation, i.e., in the case above, $\psi \left[a_{i}..b_{i}\right]=\lbrace \left[a_{i}..c_{i}\right],\left[d_{i}..b_{i}\right] \rbrace$.
If $\left|\left[a_{i}..b_{i}\right]\right|=1$, the interval
cannot be split and $\psi \left[a_{i}..b_{i}\right]=\left[a_{i}..b_{i}\right]$.
\end{definition}

@@

\begin{definition}[Splitting a region]
\label{pers02.def:Splitting-a-region}
Consider a region $R=\prod\left[a_{i}..b_{i}\right]$.The split operator $\psi$ divides it in a set of sub-regions obtained splitting each $p_i$-interval and combining the resulting sub-intervals with the cartesian product, as in \figR{split}~(b). Formally:
	\begin{align}
		\psi(R) = \left\{ \left.\prod\left[x_{i}..y_{i}\right]\right|\left[x_{i}..y_{i}\right]\in \psi\left[a_{i}..b_{i}\right]  \right\} 
	\end{align}

@@

	\begin{figure}[t]
		\begin{center}
			\subfigure[Splitting an interval.]{
				\includegraphics[width=0.30\textwidth]{img/splitting_intervals} }
			\subfigure[Splitting a region.]{
				\includegraphics[width=0.60\textwidth]{img/splitting_regions} }
		    \figLC{split}{Split operations.}
		\end{center}
	\end{figure}

@@

Examples of contiguous and non contiguous regions are given in \figR{contiguous_regions}. The two regions at the top are contiguous because their union can be expressed as the cartesian product of parameter intervals, thus they can be merged to obtain a new bigger region. The same applies also to the regions on the right. On the contrary, the two regions on the left (or of the two regions at the bottom) cannot be merged.

@@

We now describe formally the merging operations.
\begin{definition}[Merging intervals]
Let $\left[a_{i}\dots b_{i}\right]$ and $\left[x_{i}..y_{i}\right]$ be two contiguous $p_i$-intervals. The merge operation applied to them produces a new $p_i$-interval of the following form:
	\begin{align}
		\mu	 \left( \begin{array}{l}
				\left[a_{j}..y_{j}\right] \\
				\left[x_{j}..b_{j}\right]
		     \end{array} \right)
		=\begin{cases}
			\left[a_{j}..y_{j}\right] & \mbox{ if }b_{j}<x_{j}\\
			\left[x_{j}..b_{j}\right] & \mbox{ if }y_{j}<a_{j}
		\end{cases}
	\end{align}

@@

In \secR{sketch} we observed that, in order to estimate how much interesting is a region, it is crucial to understand the novelty that it introduces, i.e. how distant the Pareto-optimal configurations discovered in that region are from the Pareto-optimal configuration previously discovered. The concept of distance is expressed by the operator defined in the next definition. We prefer to call it \emph{separation} rather than \emph{distance}, because usually distance denote a commutative operator while our definition of separation is non commutative.

@@

The separation between $\mathbf{c}'$ and $\mathbf{c}''$ measures how
much we must vary the image of the former to obtain the image of the latter. Note that the separation is a \emph{normalized} measure, thanks to $o'_i$ at the denominator. This is important since usually objectives are heterogeneous, e.g. delay and power consumption, having different unity of measurement and different scale. If we did not use a normalized measure, in presence of objectives whose absolute values are small and other objectives with big values, the notion of distance would have been biased toward the latter.

>>>>

% A PROPOSITO DI QUESTO DUBBIO:
%\comment{AA: Mi sembra solo un cambio di notazione. DP: c'era un On the contrary troppo aggressivio, ho cercato di alleggerire la cosa, ma dai un'occhiata }
% HO CANCELLATO TUTTA QUESTA PARTE.
% For each point of the Pareto front we can imagine the corresponding configuration as a point of $P_{1}\times\dots\times P_{n}$ space. We will also refer to these points as ``parameter space representation'' of the Pareto front. \figR{psos} shows the relation between the objective space (containing the Pareto front) and its parameter space representation. In particular in this work we are interested in the relation between the evolution of the points in the two spaces, e.g. how some properties of the Pareto front points can reflect some features of the corresponding points in the parameter space.

\begin{figure}[t]
\figLC{psos}{Relation between Pareto-set and Pareto front.}
\center
\includegraphics[width=0.5\columnwidth]{img/Pareto_set_and_front}
\end{figure}

@@

The problem is that, in complex scenarios, not all the relations of this
kind are known in advance. Some dependency may be hidden
to the designer.
Therefore, as stated in \secR{Related-work},
setting up an exploration trying to take into account as much dependencies
as possible is a hard task that is destined to be incompletely
accomplished. We need a methodology to ``automatically'' recognize
interesting or uninteresting regions without requiring a priori knowledge
to the designer, that is what our algorithm does.

@@

\subsection{Formal definitions}
We start by defining the parameter interval and the concept of interval contiguity. Using these definition we construct the region and define the region contiguity. Then we define the operations of splitting and merging regions. Finally, we provide the definition of separation.

@@

We consider only parameters $p_{i}$ with ordering, i.e. such
that $\forall a,b\in V_{i}$ it is possible
to say $a<b$ or $a=b$ or $a>b$. 
The following definition of interval contiguity will be the base of the concept of region contiguity, used to define merging and splitting operations.
\begin{definition}[Contiguous intervals]
\label{pers02.def:Contiguous-intervals}
Two intervals are contiguous iff they do not overlap and merging them results in a new interval. More formally, let $\left[a_{i}..b_{i}\right]$ and $\left[x_{i}..y_{i}\right]$
be two $p_{i}$-intervals. They are said to be contiguous (see \figR{interval_and_region}(a) ) iff
\begin{align}\begin{array}{l}
	\left[a_{i}..b_{i}\right]\cap\left[x_{i}..y_{i}\right]=\emptyset \mbox{      and} \\
	\left[a_{i}..b_{i}\right]\cup\left[x_{i}..y_{i}\right]$ is a $p_{i}$-
interval or $\left[x_{i}..y_{i}\right]\cup\left[a_{i}..b_{i}\right]
\end{array}\end{align}
\end{definition}

@@

As already explained, we will split interesting regions and merge uninteresting ones. The region splitting and merging operations are defined starting from the corresponding operations defined on intervals.
\begin{definition}[Splitting an interval]
\defL{splitting-a-region}
Given a $p_i$ interval $\left[a_{i}..b_{i}\right]$ with more than one element, i.e. $\left|\left[a_{i}..b_{i}\right]\right|>1$, the splitting operation produces as result two contiguous intervals $\left[a_{i}..c_{i}\right],\left[d_{i}..b_{i}\right]$
 such that
	\begin{align}
		\begin{cases}
		\left[a_{i}..c_{i}\right]\cap\left[d_{i}..b_{i}\right] & =\emptyset\\
		\left[a_{i}..c_{i}\right]\cup\left[d_{i}..b_{i}\right] & =\left[a_{i}..b_{i}\right]\\
		\left|\left[a_{i}..c_{i}\right]\right| & =\left\lceil \frac{\left|\left[a_{i}..b_{i}\right]\right|}{2}\right\rceil 
		\end{cases}
	\end{align}

We define the split operator $\psi$ as the one that, when applied to the intervals, gives the result of the split operation, i.e., in the case above, $\psi \left[a_{i}..b_{i}\right]=\lbrace \left[a_{i}..c_{i}\right],\left[d_{i}..b_{i}\right] \rbrace$.
If $\left|\left[a_{i}..b_{i}\right]\right|=1$, the interval
cannot be split and $\psi \left[a_{i}..b_{i}\right]=\left[a_{i}..b_{i}\right]$.
\end{definition}

@@

\begin{definition}[Splitting a region]
\label{pers02.def:Splitting-a-region}
Consider a region $R=\prod\left[a_{i}..b_{i}\right]$.The split operator $\psi$ divides it into a set of sub-regions obtained splitting each $p_i$-interval and combining the resulting sub-intervals with the cartesian product, as shown in \figR{split}~(b). Formally:
	\begin{align}
		\psi(R) = \left\{ \left.\prod\left[x_{i}..y_{i}\right]\right|\left[x_{i}..y_{i}\right]\in \psi\left[a_{i}..b_{i}\right]  \right\} 
	\end{align}

@@

	\begin{figure}[t]
	    \figLC{split}{Split operations.}
		\begin{center}
			\subfigure[Splitting an interval.]{
				\includegraphics[width=0.30\textwidth]{img/splitting_intervals} }
			\subfigure[Splitting a region.]{
				\includegraphics[width=0.60\textwidth]{img/splitting_regions} }
		\end{center}
	\end{figure}

@@

Examples of contiguous and non contiguous regions are given in \figR{contiguous_regions}. The two regions at the top are contiguous because their union can be expressed as the cartesian product of parameter intervals, thus they can be merged to obtain a new bigger region. The same also applies to the regions on the right. On the contrary, the two regions on the left (or of the two regions at the bottom) cannot be merged.

@@

We now formally describe the merging operations.
\begin{definition}[Merging intervals]
Let $\left[a_{i}\dots b_{i}\right]$ and $\left[x_{i}..y_{i}\right]$ be two contiguous $p_i$-intervals. The merge operation applied to them produces a new $p_i$-interval of the following form:
	\begin{align}
		\mu	 \left( \begin{array}{l}
				\left[a_{j}..y_{j}\right] \\
				\left[x_{j}..b_{j}\right]
		     \end{array} \right)
		=\begin{cases}
			\left[a_{j}..y_{j}\right] & \mbox{ if }b_{j}<x_{j}\\
			\left[x_{j}..b_{j}\right] & \mbox{ if }y_{j}<a_{j}
		\end{cases}
	\end{align}

@@

In \secR{sketch} we observed that, in order to estimate how interesting a region is, it is crucial to understand the novelty it introduces, i.e. how distant the Pareto-optimal configurations discovered in that region are from the Pareto-optimal configuration previously discovered. The concept of distance is expressed by the operator defined in the next definition. We prefer to call it \emph{separation} rather than \emph{distance}, because usually distance denotes a commutative operator while our definition of separation is non commutative.

@@

The separation between $\mathbf{c}'$ and $\mathbf{c}''$ measures how
much we must vary the image of the former to obtain the image of the latter. Note that the separation is a \emph{normalized} measure, thanks to $o'_i$ at the denominator. This is important since objectives are usually heterogeneous, e.g. delay and power consumption have different unities of measurement and different scale. If we did not use a normalized measure, in presence of objectives whose absolute values are small and other objectives with big values, the notion of distance would have been biased toward the latter.
